Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-26 Paper_Draft2/Draft2.Rmd

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@equivant_response_2016, on the behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016]. @angwin2016machine maintain that there are biases in the outcome values, protected attributes, and covariates during @equivant_response_2016’s data processing phase. ProPublica collaborators @larson2016we account for these biases in their analyses. In their response, @equivant_response_2016 highlight that @angwin2016machine did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.

>>>>

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated. However, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information. @angwin2016machine has critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers racial biases in defendants’ scores [@angwin2016machine]. They find that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@equivant_response_2016, on behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016]. @angwin2016machine maintain that there are biases in the outcome values, protected attributes, and covariates during @equivant_response_2016’s data processing phase. ProPublica collaborators @larson2016we account for these biases in their analyses. In their response, @equivant_response_2016 highlight that @angwin2016machine did not account for base rates of recidivism in their analysis, which are important initial percentages without the presence of other information.
